On Determinism of Game Engines used for Simulation-based Autonomous Vehicle Verification

### 2021-02-17 paper/cav_determinisim_main.tex
half-way house, not fully completed and some parts half touched

In contrast, we argue that the precision issues related to floating-point operations are better described as \textit{incorrectness} that is in fact repeatable; they do not cause non-determinism per se. 
%
It is normally reasonable to assume that the processors on which game engines run are deterministic with respect to floating-point arithmetic. 
%
Thus, given the same operands and sequence of operations, the same processor should always produce the same bit pattern. %, i.e.\ a result that is as consistently equally \textit{incorrect} over a number of executions.
%
So, even if the result of a floating-point operation is incorrect due to floating-point rounding errors, it should always be equally \textit{incorrect} for implementations that conform to the IEEE floating-point standard~\cite{8766229}.

@@

\subsection{NUMA Non-Uniform Memory Access}
For a repeated test that operates over a number of cores based on a CPU scheduling policy, memory access time may vary depending on the physical memory location relative to the processor. Typically a core can access its own memory with lower latency than that of another core resulting in lower interprocessor data transfer cost~\cite{nieplocha1996global}. 
%
Changes in latency between repeated tests may, in the worst case, cause the game engine to operate non-deterministically if tasks are processed out of sequence using equal priority scheduling, or, perhaps, simply with an increased data transfer cost, i.e.\ slower. 
%
By binding a process to a specific core for the duration of its execution, the variations in data transfer time can be minimised.

@@

\subsection{Summary}
We have investigated the potential sources of non-determinism affecting game engines 
and explored the impact they may have on simulation variance. 
%
% \todo[inline,color=red!40]{GC need to decide if fp is a source or not: Assuming deterministic hardware, there is no contribution to simulation variance due to floating-point representation and any rounding errors due to bit width limits should be repeatable. }
%
Memory checking not withstanding, errors associated with the lack of ECC are likely to be minimal unless there is significant background radiation or 1000's of hours of computation are expected.
\todo[inline]{So, do commercial HPC facilities have ECC memory?}
\todo[inline,color=green!40]{GC yes see Dell1997 "Today all servers except the very low end support Single Error Correct (SEC) ECC"}
%
\todo[inline,color=red!40]{old: To ensure repeatable simulation outcomes the physics setting $dt$ must be fixed, along with any actor navigation meshes, seeds for random number generation, game engine setup and simulation specific parameters.}
To ensure precise simulation outcomes the physics setting, $dt$, must be fixed, along with any actor navigation meshes, seeds for random number generation, game engine setup and simulation specific parameters.
\todo[inline,color=green!40]{GC alternative: To ensure precise simulation outcomes the physics setting $dt$ must be fixed, along with any actor navigation meshes, random number seeds, game engine setup and simulation specific parameters.}
%
NUMA should only affect interprocessor data transfer cost and, without control measures, will only make the computation cycle longer. Relative access times between different caches are likely to be small although may have a more pronounced impact on high throughput systems, e.g.\ HPC. {\color{red} Not sure this is correct, see the NUMA section above. Why ``only''? If one computation out of many takes longer, then their exec order can be affected, or not?}
%
Basic thread scheduling should not affect the simulation's determinism unless changing scheduling policy, operating system or migrating between machines with different setups. 
%
However, should new and unexpected threads start during the simulation, then the interruption to execution order or additional resource demand may affect timing of subsequent steps, thus reducing the number of physics updates within a game loop. Likewise, uncontrolled allocation of hardware resources such as CPUs or GPUs can potentially give rise to non-determinism. 
%

@@

\todo[inline,color=red!40]{old:  We first describe the context, scene and scenario of interest before discussing and defining a threshold for what is considered an acceptable simulation variance in this context.}
%
We first describe the context, scene and scenario of interest before discussing and defining a tolerance for what is considered an acceptable simulation variance in this context.
%
\todo[inline,color=green!40]{GC alternative: We first describe the context, scene and scenario of interest before discussing and defining a tolerance for what is considered an acceptable simulation variance in this context.}
%
\todo[inline]{Complete the section structure as this is a very long section.}

@@

@@

After discovering that internal engine settings could not achieve a simulation variance that met the required 
% \textit{deterministic threshold}, 
\textit{tolerance}, {\color{red} WHEN HAS THIS BEEN REPORTED?}
settings external to the game engine were explored. 
%
Game engines, and simulation hardware more generally, will utilise available system resources such as central and graphical processing units (CPU and GPU respectively), to perform physics calculations and run the simulation environment. 
%
For high performance simulations the demand on these resources may be a significant fraction of the total available and an initial hypothesis was that as this ratio tended toward one, there would be an increase in \textit{simulation variance}. This hypothesis was supported by our initial work performed on the unreal engine~\cite{TSLUnrealEngineTesting} and was explored more fully in this work using the CARLA platform. {\color{red} From a reviewer's point of view this is a mysterious initial study that has been referred to often but nothing has been presented. Either take out or include, at least wrt.\ the overall observations, such as the support for this hypothesis.}

To replicate in a controlled manner the high computational loads that may be anticipated for high performance simulations, software that artificially utilises resources on both the CPU and GPU were executed alongside the simulation. 
%
% In order to reproduce conditions for computationally intensive simulations, the experiments are executed with additional levels of artificial resource utilisation. 
Resource utilisation was artificially increased for both CPU and GPU devices to include a range of values from 0 to 95\% (see Section~\ref{s:methodology}) using reported values of the system monitors \texttt{htop} and \texttt{nvidia-smi} respectively. Resource utilisation figures reported here should be considered approximate values.
% The level of additional resource utilisation, above that of the needs of the simulation, was increased from $0\%$ (no additional resource load) in $25\%$ increments up to a maximum of $95\%$ for both CPU and GPU as reported by system monitors (\texttt{htop} and \texttt{nvidia-smi} respectively).

@@

% ***************************************************
%  System Configuration and Screening
% ***************************************************
\subsection{System Configuration and Screening}\label{s:screening}
These tests were carried out on an Alienware Area 51 R5 with an i9 9960X processor with 64GB non-ECC DDR4 RAM at 2933MHz with an NVIDIA GeForce RTX 2080 GPU with 8GB GDDR6 RAM at 1515 MHz. 
%
Operating systems was Linux Ubuntu 18.04.4 LTS. To ensure reliability of results, each test was repeated 1000 times. 
%
Tests were carried out in CARLA (v0.9.6 and Unreal Engine v4.22) using synchronous mode with a fixed $dt$ of $0.05$s. 
%
A detailed guide for reproducing these experiments along with the scripts used are provided on github\footnote{\url{https://github.com/TSL-UOB/CAV-Determinism/tree/master/CARLA_Tests_setup_guide}}.
%
To eliminate some of the potential sources of non-determinism outlined in Section~\ref{s:nondeterminisimSources} a series of screening tests and analyses were performed on our system. These were:

\begin{itemize}[leftmargin=*]
    \item System memory: \texttt{memtest86}~\cite{MemTest86} full suite of tests ran, all passed.
    \item Graphical memory: \texttt{cuda\_memtest}~\cite{cuda_memtest}\cite{shi2009testing}.
    \item Non-uniform memory access: \texttt{numactl}~\cite{numactl_NUMA} was used to fix the simulator and test script to single cores but only a minor (2\%) improvement in simulation variance was observed and therefore not used for subsequent testing.
\end{itemize}

@@

\begin{figure*}[t]
    \centering
    \begin{subfigure}{.49\textwidth}
        \includegraphics[width=1\textwidth]{Other/Figures/CarsCollisionCG25_V4.pdf}
        \caption{}
        \label{CarsCollisionCG25}
    \end{subfigure}
    \begin{subfigure}{.49\textwidth}
        \includegraphics[width=1\textwidth]{Other/Figures/CarsCollisionPrePostV5.pdf}
        \caption{}
        \label{CarsCollisionPrePost}
    \end{subfigure}
    \caption{Vehicle to vehicle collision (test 2) showing (a) maximum deviation against simulation time for 25\% resource utilisation and (b) maximum deviation pre- and post-collision against resource utilisation. The simulation noise floor is shown in (a) which is the empirical lower limit of deviation for the hardware reported in this study.}
\end{figure*}

% ***************************************************
%  Results and Discussion
% ***************************************************
\subsection{Results and Discussion}\label{ResultsSection}
A summary of all the results are shown Table~\ref{TableOfExperiments} in the column $\max\sigma$ (unrestricted), where the value reported is the maximum deviation across all resource utilisation levels, i.e. the worst case for a given scenario.  
%
From these  results it is clear that scenarios with only pedestrian actors (tests 5-6) display results within tolerance over all resource utilisation levels with or without a collision where $\max\sigma$ is $5.6\times10^{-13}$ or $0.56\si{\pico\metre}$. However, all other scenarios involving vehicles or a mixture of actor types breach the tolerance that has been set, with some deviation in actor path as large as $59$cm.
\todo[inline,color=green!40]{GC Edit: From these  results it is clear that scenarios with only pedestrian actors (tests 5-6) display results within the permissible tolerance over all resource utilisation levels with or without a collision where $\max\sigma$ is $5.6\times10^{-13}$ or $0.56\si{\pico\metre}$. However, all other scenarios involving vehicles or a mixture of actor types breach the permissible tolerance that has been set, with some deviation in actor path as large as $59$cm. }
%
Clearly a deviation in results of such a large amount is unacceptable if simulation is to be used as a reliable verification tool.

Resource utilisation level was found to have a significant impact on \textit{simulation variance}.
%
Figure~\ref{ExperimentsStressSummary} shows $\max\sigma$ against the artificially increased resource utilisation level, where the $x$-axis indicates the approximate percentage of resource utilisation (for CPU \& GPU). In this figure, anything above the $1$cm level (dashed line) is considered non-permissible according to our specific tolerance level informed from our verification requirements.
%
\todo[inline]{Where are the verification requirements listed? They should be part of the scenario description. GC- added to the start of section 3.} 
%

A general pattern in the results indicates that; some scenarios consistently fail to produce results within permissible levels of variance at any resource utilisation level (Fig.~\ref{ExperimentsStressSummary} Test ID 2 \& 4 above the dashed cm level), some are consistently within tolerance (Test ID 5 \& 6 with pedestrians only), and that some cases only fail to meet the tolerance requirement when at higher utilisation levels, breaching the tolerance above 75\% resource utilisation (Test ID 1 \& 3).

\todo[inline]{Help the reader see what you can see by explaining exactly where this can be seen in the figure(s). GC text added above}

However, the non-permissible results in Figure~\ref{ExperimentsStressSummary} (all those above the dashed line) are the worst case account of the situation, as per Equation~\ref{eq:max_sigma}, as the maximum variance is taken over the entire simulation period.

Examining specifically the results from tests 2 \& 4 as a function of simulation time reveals further information about the simulation variance before and after an actor collision. Fig.~\ref{CarsCollisionCG25} shows this examination for vehicle to vehicle collisions (test 2), where $\max\sigma$ switches from permissible prior to the vehicle collision to non-permissible post-collision. 
%
This pattern of permissible results prior to collision and non-permissible post-collision is maintained up to a resource utilisation level of approximately 75\%, see Fig.~\ref{CarsCollisionPrePost}. This time series examination was repeated for vehicle to pedestrian collisions (test 4) and the results are shown in Fig.~\ref{CarsPeopleCollsionsCG25}. Similarly to vehicle-to-vehicle collisions, the variation of $\max\sigma$ for vehicle to pedestrian collisions indicates permissible pre-collision behaviour with up to 75\% resource utilisation, see Fig.~\ref{CarsPeopleCollisionPrePost}. This is a key finding of this work; it suggests that verification engineers should consider terminating tests at the point of a collision, as any post-collision results will be non-permissible.
\todo[inline,color=green!40]{GC Edit: Examining specifically the results from tests 2 \& 4 as a function of simulation time reveals further information about the simulation variance before and after an actor collision. Fig.~\ref{CarsCollisionCG25} shows this examination for vehicle to vehicle collisions (test 2), where $\max\sigma$ switches from permissible prior to the vehicle collision to non-permissible post-collision. 
%
This pattern of permissible results prior to collision and non-permissible post-collision is maintained up to a resource utilisation level of approximately 75\%, see Fig.~\ref{CarsCollisionPrePost}. This time series examination was repeated for vehicle to pedestrian collisions (test 4) and the results are shown in Fig.~\ref{CarsPeopleCollsionsCG25}. Similarly to vehicle-to-vehicle collisions, the variation of $\max\sigma$ for vehicle to pedestrian collisions indicates permissible pre-collision behaviour with up to 75\% resource utilisation, see Fig.~\ref{CarsPeopleCollisionPrePost}. This is a key finding of this work; it suggests that verification engineers should consider terminating tests at the point of a collision, as any post-collision results will be non-permissible. }
% \todo{Why ``not reliable''.}

\begin{figure*}[h]
    \centering
    \begin{subfigure}{.49\textwidth}
        \includegraphics[width=1\textwidth]{Other/Figures/CarsPeopleCollsionsCG25_V3.pdf}
        \caption{}
        \label{CarsPeopleCollsionsCG25}
    \end{subfigure}
    \begin{subfigure}{.49\textwidth}
        \includegraphics[width=1\textwidth]{Other/Figures/CarsPeopleCollisionPrePostV5.pdf}
        \caption{}
        \label{CarsPeopleCollisionPrePost}
    \end{subfigure}
    \caption{Vehicle to pedestrian collision (test 4) showing (a) maximum deviation against simulation time for 25\% resource utilisation and (b) maximum deviation pre- and post-collision for different resource utilisation levels.}
\end{figure*}

\begin{figure}[t]
    \centering
    \includegraphics[width=0.48\textwidth]{Other/Figures/NICE_analysis_summary_V4.pdf}
    \caption{Summary of investigating the effect of NICE priority setting for 0\% and 75\% additonal CPU \& GPU resource utilisation.}
    \label{NICEExperimentStressSummary}
\end{figure}

The second key finding of this work is illustrated in Fig.~\ref{CarsPeopleCollsionsCG25}. In this scenario (test 4), there is a collision between a vehicle (Car ID 2, solid line) and a pedestrian (Ped ID 3, dot dash line) which occurs at a simulation time of approximately $6$s and a second vehicle actor (Car ID 1, dashed line), which is \textit{not involved in the collision}. There are three observations here; firstly that the vehicle directly involved in the collision (Car ID 2) displays high simulation variance immediately after the collision.
% as described above. 
Secondly, that the maximum deviation of the pedestrian involved in the collision (Ped ID 3) is at a tolerable level throughout the test\footnote{However, please note that in CARLA the pedestrian object is destroyed post-collision hence the flat line from $t=6$s onwards.}. Thirdly, we observed a delayed effect on Car ID 1 showing high simulation variance with a $5$s delay \textit{even though this vehicle was not involved in the collision}. This final point should be of particular concern to verification engineers and research practitioners in the field as it implies that \textit{any collision between actors can affect the simulation variance of the entire actor population} and could potentially result in  erroneous assertion checking.

@@

% This was calculated by averaging the variance between all of the different runs through out the simulation time for all of the agents, then taking the square root to obtain the total mean devaiton. Note the dotted red line in the plot shows the millimeter (mm) level. Below this level we assume that the simulation variance is low enough that it would not affect our assertion checking for V\&V of AV simulations purposes, that it is sufficient to assume the tests are deterministic and repeatable. This line can be moved up or down and it is entirely up to verification engineers where they want to define the level below which tests' variance is negligibale and it is suffcient to assume that tests are deterministic.
%
\begin{figure*}[t]
    \centering
    \includegraphics[width=0.99\linewidth]{Other/Figures/Methodology_Diagram_v5.pdf}
    \caption{Flow diagram of the methodology proposed to determine the simulation variance of a simulation.}
    \label{method_diagram}
\end{figure*}

@@

The main findings of this work suggest a working practice that would minimise the non-deterministic effects observed in this investigation. By limiting simulation results to pre-collision data and ensuring resource utilisation levels do not exceed 75\%, the permissible variance of $1$cm is achievable as shown in the \textit{restricted} column in Table.~\ref{TableOfExperiments}. By applying this set of restrictions upon the simulation the maximum observed deviation across all experiments was $0.98$cm which is within the target tolerance we set out to achieve. 
%
Practitioners may wish to set a stricter resource utilisation level, such as less than 50\% to further reduce the potential dispersion of results if this is required for their chosen application. 
%
\todo[inline,color=green!40]{GC Addition: Practitioners may wish to set a stricter resource utilisation level, such as less than 50\% to further reduce the potential dispersion of results if this is required for their chosen application. 
%For this case study a 50\% restriction of resource utilisation would reduce deviation across all experiments to XXX.XXXmm. 
}

\subsection{Process Scheduling Priority}
An investigation into the response of process scheduling on the simulation variance was undertaken. The experiment was repeated ($n=1000$) using test 1 but altering the process scheduling priority using the program \texttt{NICE}.\footnote{\url{http://manpages.ubuntu.com/manpages/bionic/man1/nice.1.html}} Setting a higher priority for the simulator process with respect to the resource utilisation processes, it was possible to determine if scheduling could account for the increased simulation variance when the system is under high resource utilisation.
%
To give a process a high priority a negative NICE value is set with the lowest being -20. To decrease the priority a positive value is set, up to +19. The default NICE value is 0.

The results are presented in Fig.~\ref{NICEExperimentStressSummary} where the box denotes the inter-quartile range, non-outlier limits by the whiskers and a horizontal bar for the median. The figure shows that decreasing the priority (right hand side of plot) has little effect on simulation variance when compared to a NICE value of 0 (central plot bars). Increasing priority (left hand side of plot) reduced variance for the 75\% resource utilisation but this did not account for all the difference in the observed results. This is demonstrated in the maximum priority setting where the bars in plot are not equal, indicating an additional contribution to variance not accounted for by the NICE scheduling. 
%
This unaccounted difference in the variance may be due to the lack of absolute control that \texttt{NICE} has over the process scheduling.\footnote{\url{https://askubuntu.com/questions/656771/process-niceness-vs-priority}}

\subsection{Investigation Summary} \label{s:empirical_summary}

>>>>

In contrast, we argue that the precision issues related to floating-point operations are better described as \textit{incorrectness} that is in fact repeatable; they do not cause non-determinism per se. {\color{red} Should this be IMPRECISION rather than incorrectness?}
%
It is normally reasonable to assume that the processors on which game engines run are deterministic with respect to floating-point arithmetic. 
%
Thus, given the same operands and sequence of operations, the same processor should always produce the same bit pattern. %, i.e.\ a result that is as consistently equally \textit{incorrect} over a number of executions.
%
So, even if the result of a floating-point operation is incorrect due to floating-point rounding errors, it should always be equally \textit{incorrect} for implementations that conform to the IEEE floating-point standard~\cite{8766229}.

@@

\subsection{Non-Uniform Memory Access (NUMA)}
For a repeated test that operates over a number of cores based on a CPU scheduling policy, memory access time may vary depending on the physical memory location relative to the processor. Typically a core can access its own memory with lower latency than that of another core resulting in lower interprocessor data transfer cost~\cite{nieplocha1996global}. 
%
Changes in latency between repeated tests may, in the worst case, cause the game engine to operate non-deterministically if tasks are processed out of sequence using equal priority scheduling, or, perhaps, simply with an increased data transfer cost, i.e.\ slower. 
%
By binding a process to a specific core for the duration of its execution, the variations in data transfer time can be minimised.

@@

\subsection{Summary}
We have investigated the potential sources of non-determinism affecting game engines 
and explored the impact they may have on simulation variance. 
%
% \todo[inline,color=red!40]{GC need to decide if fp is a source or not: Assuming deterministic hardware, there is no contribution to simulation variance due to floating-point representation and any rounding errors due to bit width limits should be repeatable. }
%
Memory checking not withstanding, errors associated with the lack of ECC are likely to be minimal unless there is significant background radiation or 1000's of hours of computation are expected.
\todo[inline]{So, do commercial HPC facilities have ECC memory?}
\todo[inline,color=green!40]{GC yes see Dell1997 "Today all servers except the very low end support Single Error Correct (SEC) ECC"}
%
\todo[inline,color=red!40]{old: To ensure repeatable simulation outcomes the physics setting $dt$ must be fixed, along with any actor navigation meshes, seeds for random number generation, game engine setup and simulation specific parameters.}
To ensure precise simulation outcomes the physics setting, $dt$, must be fixed, along with any actor navigation meshes, seeds for random number generation, game engine setup and simulation specific parameters.
\todo[inline,color=green!40]{GC alternative: To ensure precise simulation outcomes the physics setting $dt$ must be fixed, along with any actor navigation meshes, random number seeds, game engine setup and simulation specific parameters.}
%
Non-uniform memory access (NUMA) should only affect interprocessor data transfer cost and, without control measures, will only make the computation cycle longer. Relative access times between different caches are likely to be small although may have a more pronounced impact on high throughput systems, e.g.\ HPC. {\color{red} Not sure this is correct, see the NUMA section above. Why ``only''? If one computation out of many takes longer, then their exec order can be affected, or not?}
%
Basic thread scheduling should not affect the simulation's determinism unless changing scheduling policy, operating system or migrating between machines with different setups. 
%
However, should new and unexpected threads start during the simulation, then the interruption to execution order or additional resource demand may affect timing of subsequent steps, thus reducing the number of physics updates within a game loop. Likewise, uncontrolled allocation of hardware resources such as CPUs or GPUs can potentially give rise to non-determinism. 
%

@@

\todo[inline,color=red!40]{old:  We first describe the context, scene and scenario of interest before discussing and defining a threshold for what is considered an acceptable simulation variance in this context.}
%
We first describe the context, scene and scenario of interest before discussing and defining a tolerance for what is considered an acceptable simulation variance in this context. {\color{red} and then..? Please cover the entire section - consider splitting setup from evaluation.}
%
\todo[inline,color=green!40]{GC alternative: We first describe the context, scene and scenario of interest before discussing and defining a tolerance for what is considered an acceptable simulation variance in this context.}
%
\todo[inline]{Complete the section structure as this is a very long section.}

@@

\subsection{Initial Results and Discussion}\label{ResultsSection}

{\color{red} This is for unrestricted, so this does not come as a surprise later.}

@@

After discovering that internal engine settings could not achieve a simulation variance that met the required 
% \textit{deterministic threshold}, 
\textit{tolerance}, {\color{red} WHEN HAS THIS BEEN REPORTED? Should there be an ``Interim Results'' section?}
settings external to the game engine were explored. 
%
Game engines, and simulation hardware more generally, will utilise available system resources such as central and graphical processing units (CPU and GPU respectively), to perform physics calculations and run the simulation environment. 
%
For high performance simulations the demand on these resources may be a significant fraction of the total available and an initial hypothesis was that as this ratio tended toward one, there would be an increase in \textit{simulation variance}. This hypothesis was supported by our initial work performed on the unreal engine~\cite{TSLUnrealEngineTesting} and was explored more fully in this work using the CARLA platform. {\color{red} From a reviewer's point of view this is a mysterious initial study that has been referred to often but nothing has been presented. Either take out or include, at least wrt.\ the overall observations, such as the support for this hypothesis.}

To replicate in a controlled manner the high computational loads that may be anticipated for high performance simulations, software that artificially utilises resources on both the CPU and GPU were executed alongside the simulation. 
%
% In order to reproduce conditions for computationally intensive simulations, the experiments are executed with additional levels of artificial resource utilisation. 
Resource utilisation was artificially increased for both CPU and GPU devices to include a range of values from 0 to 95\% (see Section~\ref{s:methodology}) using reported values of the system monitors \texttt{htop} and \texttt{nvidia-smi} respectively. Resource utilisation figures reported here should be considered approximate values.
% The level of additional resource utilisation, above that of the needs of the simulation, was increased from $0\%$ (no additional resource load) in $25\%$ increments up to a maximum of $95\%$ for both CPU and GPU as reported by system monitors (\texttt{htop} and \texttt{nvidia-smi} respectively). 
%
{\color{red} Somwhere the experimental setup needs to clarify that one set was executed ``unrestricted'' and another ``restricted'' - say what this means and why this was done.}

@@

% ***************************************************
%  System Configuration and Screening
% ***************************************************
\subsection{Experimental System Configuration and Pre-Screening}\label{s:screening}
The experiments were carried out on an Alienware Area 51 R5 with an i9 9960X processor with 64GB non-ECC DDR4 RAM at 2933MHz with an NVIDIA GeForce RTX 2080 GPU with 8GB GDDR6 RAM at 1515 MHz. 
%
The operating systems was Linux Ubuntu 18.04.4 LTS. 
%
Tests were carried out in CARLA (v0.9.6 and Unreal Engine v4.22) using synchronous mode with a fixed $dt$ of $0.05$s. 
%
Each test was repeated 1000 times. 
%
A detailed guide for reproducing the experiments along with the scripts used are provided on github\footnote{\url{https://github.com/TSL-UOB/CAV-Determinism/tree/master/CARLA_Tests_setup_guide}}.
%
To eliminate some of the potential sources of non-determinism outlined in Section~\ref{s:nondeterminisimSources} a series of screening tests and analyses were performed on our system. These were:

\begin{itemize}[leftmargin=*]
    \item System memory: \texttt{memtest86}~\cite{MemTest86} full suite of tests ran, all passed.
    \item Graphical memory: \texttt{cuda\_memtest}~\cite{cuda_memtest}, \cite{shi2009testing}.
    \item Non-uniform memory access (NUMA): \texttt{numactl}~\cite{numactl_NUMA} was used to fix the simulator and test script to single cores but only a minor (2\%) improvement in simulation variance was observed. Therefore this was not used for subsequent testing. {\color{red} But if we aim to min the sources of non-det, then that should be considered/used?}
\end{itemize}

@@

\begin{figure*}[t]
    \centering
    \begin{subfigure}{.49\textwidth}
        \includegraphics[width=1\textwidth]{Other/Figures/CarsCollisionCG25_V4.pdf}
        \caption{}
        \label{CarsCollisionCG25}
    \end{subfigure}
    \begin{subfigure}{.49\textwidth}
        \includegraphics[width=1\textwidth]{Other/Figures/CarsCollisionPrePostV5.pdf}
        \caption{}
        \label{CarsCollisionPrePost}
    \end{subfigure}
    \caption{Vehicle to vehicle collision (Test 2) showing (a) maximum deviation against simulation time for 25\% resource utilisation and (b) maximum deviation pre- and post-collision against resource utilisation. The simulation noise floor is shown in (a) which is the empirical lower limit of deviation for the hardware reported in this study.}
\end{figure*}

% ***************************************************
%  Results and Discussion
% ***************************************************
\subsection{Final Results and Discussion}\label{ResultsSection}

{\color{red} Consider spitting to introduce the first part of the table, which motivates the 2nd lot of experiments under load restrictions. I'd find this more logical.}

A summary of all the results is shown in Table~\ref{TableOfExperiments}.  In the column $\max\sigma$ (unrestricted) the value reported is the maximum deviation across all resource utilisation levels, i.e.\ the worst case for a given scenario.  
%
From these results it is clear that scenarios with only pedestrian actors (Tests 5 \& 6) display results within tolerance over all resource utilisation levels with or without a collision where $\max\sigma$ is $5.6\times10^{-13}$ or $0.56\si{\pico\metre}$. 
%
However, all other scenarios involving vehicles or a mixture of actor types do not meet the required tolerance, with some deviation in actor path as large as $59$cm.
%
\todo[inline,color=green!40]{GC Edit: From these  results it is clear that scenarios with only pedestrian actors (tests 5-6) display results within the permissible tolerance over all resource utilisation levels with or without a collision where $\max\sigma$ is $5.6\times10^{-13}$ or $0.56\si{\pico\metre}$. However, all other scenarios involving vehicles or a mixture of actor types breach the permissible tolerance that has been set, with some deviation in actor path as large as $59$cm. }
%
Clearly, such a large deviation cannot be acceptable for simulation to be considered a credible verification tool.

Resource utilisation was found to have a significant impact on \textit{simulation variance}.
%
Figure~\ref{ExperimentsStressSummary} shows $\max\sigma$ against the artificially increased resource utilisation level, where the $x$-axis indicates the approximate percentage of resource utilisation (for CPU \& GPU). In this figure, any $\max\sigma$ above the $1$cm level (indicated by a dashed line) is considered non-permissible according to our specified tolerance. 
%
Note that the non-permissible results in Figure~\ref{ExperimentsStressSummary} (all those above the dashed line) are the worst case account of the situation, as per Equation~\ref{eq:max_sigma}, as the maximum variance is taken over the entire simulation period.
%
\todo[inline]{Where are the verification requirements listed? They should be part of the scenario description. GC- added to the start of section 3.} 
%

A general pattern in the results indicates that some scenarios consistently fail to produce results within tolerance, irrespective of resource utilisation (cf.\ Fig.~\ref{ExperimentsStressSummary} Test 2 \& 4 are above the dashed $1$cm line), while some are consistently within tolerance (cf.\ Fig.~\ref{ExperimentsStressSummary} Test 5 \& 6 both are with pedestrians only), and some cases only fail to meet the tolerance requirement at higher resource utilisation levels, i.e.\ above 75\% resource utilisation (cf.\ Fig.~\ref{ExperimentsStressSummary} Test 1 \& 3). {\color{red} ??? about 85\% based on the figure - or not???}

Examining specifically the results from Tests 2 \& 4 as a function of simulation time reveals further information about the simulation variance before and after an actor collision. Fig.~\ref{CarsCollisionCG25} shows this examination for vehicle to vehicle collisions (Test 2), where $\max\sigma$ switches from permissible prior to the vehicle collision to non-permissible post collision. 
%
The pattern of permissible results prior to collision and non-permissible post collision is maintained up to a resource utilisation level of approximately 75\% {\color{red} Why 75\%?}, see Fig.~\ref{CarsCollisionPrePost}. This time series examination was repeated for vehicle to pedestrian collisions (Test 4) and the results are shown in Fig.~\ref{CarsPeopleCollsionsCG25}. Similarly to vehicle-to-vehicle collisions, the variation of $\max\sigma$ for vehicle to pedestrian collisions indicates permissible pre-collision behaviour with up to 75\% resource utilisation, see Fig.~\ref{CarsPeopleCollisionPrePost}. This is a key finding; it suggests that verification engineers should consider terminating tests at the point of a collision, as any post-collision results will be non-permissible.

\todo[inline,color=green!40]{GC Edit: Examining specifically the results from tests 2 \& 4 as a function of simulation time reveals further information about the simulation variance before and after an actor collision. Fig.~\ref{CarsCollisionCG25} shows this examination for vehicle to vehicle collisions (test 2), where $\max\sigma$ switches from permissible prior to the vehicle collision to non-permissible post-collision. 
%
This pattern of permissible results prior to collision and non-permissible post-collision is maintained up to a resource utilisation level of approximately 75\%, see Fig.~\ref{CarsCollisionPrePost}. This time series examination was repeated for vehicle to pedestrian collisions (test 4) and the results are shown in Fig.~\ref{CarsPeopleCollsionsCG25}. Similarly to vehicle-to-vehicle collisions, the variation of $\max\sigma$ for vehicle to pedestrian collisions indicates permissible pre-collision behaviour with up to 75\% resource utilisation, see Fig.~\ref{CarsPeopleCollisionPrePost}. This is a key finding of this work; it suggests that verification engineers should consider terminating tests at the point of a collision, as any post-collision results will be non-permissible. }
% \todo{Why ``not reliable''.}

\begin{figure*}[h]
    \centering
    \begin{subfigure}{.49\textwidth}
        \includegraphics[width=1\textwidth]{Other/Figures/CarsPeopleCollsionsCG25_V3.pdf}
        \caption{}
        \label{CarsPeopleCollsionsCG25}
    \end{subfigure}
    \begin{subfigure}{.49\textwidth}
        \includegraphics[width=1\textwidth]{Other/Figures/CarsPeopleCollisionPrePostV5.pdf}
        \caption{}
        \label{CarsPeopleCollisionPrePost}
    \end{subfigure}
    \caption{Vehicle to pedestrian collision (Test 4) showing (a) maximum deviation against simulation time for 25\% resource utilisation and (b) maximum deviation pre- and post-collision for different resource utilisation levels.}
\end{figure*}

\begin{figure}[t]
    \centering
    \includegraphics[width=0.48\textwidth]{Other/Figures/NICE_analysis_summary_V4.pdf}
    \caption{Variance range of three NICE priority settings for additional CPU \& GPU resource utilisation of 0\% and 75\%.}
    \label{NICEExperimentStressSummary}
\end{figure}

The second key finding of this work is illustrated in Fig.~\ref{CarsPeopleCollsionsCG25}. In this scenario (Test 4), there is a collision between a vehicle (Car ID 2, solid line) and a pedestrian (Ped ID 3, dot dash line) which occurs at a simulation time of approximately $6$s and a second vehicle actor (Car ID 1, dashed line), which is \textit{not involved in the collision}. There are three observations; firstly that the vehicle directly involved in the collision (Car ID 2) displays high simulation variance immediately after the collision.
% as described above. 
Secondly, that the maximum deviation of the pedestrian involved in the collision (Ped ID 3) is at a tolerable level throughout the test\footnote{However, please note that in CARLA the pedestrian object is destroyed post-collision hence the flat line from $t=6$s onwards.}. Thirdly, we observed a delayed effect on Car ID 1 showing high simulation variance with a $5$s delay \textit{even though this vehicle was not involved in the collision}. This final point should be of particular concern to verification engineers, developers and researchers in the field as it implies that \textit{any collision between actors can affect the simulation variance of the entire actor population} and could potentially result in  erroneous simulation results. % not just assertion checking

@@

% This was calculated by averaging the variance between all of the different runs through out the simulation time for all of the agents, then taking the square root to obtain the total mean devaiton. Note the dotted red line in the plot shows the millimeter (mm) level. Below this level we assume that the simulation variance is low enough that it would not affect our assertion checking for V\&V of AV simulations purposes, that it is sufficient to assume the tests are deterministic and repeatable. This line can be moved up or down and it is entirely up to verification engineers where they want to define the level below which tests' variance is negligibale and it is suffcient to assume that tests are deterministic.
%
\begin{figure*}[t]
    \centering
    \includegraphics[width=0.99\linewidth]{Other/Figures/Methodology_Diagram_v5.pdf}
    \caption{Stages of the methodology proposed to determine the variance of a simulation.}
    \label{method_diagram}
\end{figure*}

@@

To conclude, the main findings of this case study suggest a working practice that would minimise the factors that give rise to the non-deterministic effects observed in this investigation. 
%
By limiting simulation results to pre-collision data and ensuring resource utilisation levels do not exceed 75\%, the permissible variance of $1$cm is achievable as shown in the \textit{restricted} column in Table~\ref{TableOfExperiments}. By applying this set of restrictions upon the simulation the maximum observed deviation across all experiments was $0.98$cm which is within the target tolerance we set out to achieve. 
%
Practitioners may wish to set a stricter resource utilisation level, such as less than 50\% to further reduce the potential dispersion of results if this is required for their chosen application. 
%
\todo[inline,color=green!40]{GC Addition: Practitioners may wish to set a stricter resource utilisation level, such as less than 50\% to further reduce the potential dispersion of results if this is required for their chosen application. 
%For this case study a 50\% restriction of resource utilisation would reduce deviation across all experiments to XXX.XXXmm. 
}

\subsection{Process Scheduling Priority}
An investigation into the impact of process scheduling on the simulation variance was undertaken {\color{red} say why}. The experiment was repeated ($n=1000$) using Test 1 but altering the process scheduling priority using the program \texttt{NICE}.\footnote{\url{http://manpages.ubuntu.com/manpages/bionic/man1/nice.1.html}} Setting a higher priority for the simulator process with respect to the resource utilisation processes, it was possible to determine if scheduling could account for the increased simulation variance when the system is under high resource utilisation.
%
To give a process a high priority a negative NICE value is set with the lowest being -20. To decrease the priority a positive value is set, up to +19. The default NICE value is 0.
